Improved state-space models for inference about spatial and temporal variation in abundance from count data

### 2012-09-28 Dail_Mad_ext.tex
original formulation

As originally formulated, the DM model includes submodels for three
conditionally related processes: (1) initial abundance, i.e. the
abundance at site $i$ during the first primary period,
denoted $[N_{i1}]$, (2) subsequent abundance, which is dependent upon
abundance in previous years $[N_{it}|N_{it-1}]$, and (3) the
detection process, $[n_{it}|N_{it}]$ \citep{dail_madsen:2011}.
The first two processes describe the state process---the
variation in abundance in space and time. The third item in the list
is the observation process, which describes the relationship between
the true state, abundance, and the observed count data.

@@

In the original formulation, initial abundance was
modeled using either the Poisson or negative binomial distributions:
\begin{gather}
N_{i1} \sim \mathrm{Pois}(\Lambda_i) \nonumber \\
\text{or} \nonumber \\
N_{i1} \sim \mathrm{NB}(\Lambda_i, \alpha)
\label{eq:N1}
\end{gather}
where
%$N_{i1}$ is the abundance at site $i=1,\hdots,R$ during year 1 and
$\Lambda_i$ is the expected abundance at site $i$ during year 1. The
Poisson distribution assumes that the mean of $N_{i1}$ is equal to its
variance, whereas the negative binomial distribution allows the
variance to be greater than the mean with the amount of overdispersion
determined by the parameter $\alpha$.

Regardless of the specified distribution, the model for initial
abundance has two distinguising features. First, it provides a
mechanism for characterizing spatial variation in abundance. For
instance, one might consider the influence of some environmental
covariate ($x_i$) on abundance using a log-linear
model such as $\log(\Lambda_i) = \beta^{\Lambda}_0 + \beta^{\Lambda}_1
x_{i}$. The second important point is that the spatial replicates resolve the
problem of parameter non-identifiability that are common to standard
state-space models because, as demonstrated by \citet{royle:2004biom},
process variation and observation error can be estimated from
spatially-replicated count data. Hence, the first component of the
model addresses issues (1) and (2) identified in the introduction.

@@

The DM model assumes that abundance in time $t$ is a function of
abundance in time $t-1$, i.e. abundance at each site evolves as a
first order Markovian process, although
higher order processes are also possible.  \citet{dail_madsen:2011}
considered several models to describe the temporal dynamics; however,
in each case they modeled $N_t$ as the sum of two random variables:
$S_t$, the number of indivduals surviving from $t-1$ and not
emigrating; and $G_t$ the number of new individuals entering the
population. Their most general model was
\begin{equation}
\left.\begin{aligned}
S_{it}|N_{it-1} &\sim \mathrm{Bin}(N_{it-1}, \omega) \\
G_{it}|N_{it-1} &\sim \mathrm{Pois}(\gamma(N_{it-1})) \\
N_{it} &= S_{it}+G_{it}
\end{aligned}\right\} \quad \text{for} \; t=2,\hdots,T
\label{eq:Nt}
\end{equation}
where $\omega$ is the apparent survival probability and $\gamma$ is
the recruitment rate (which can depend on $N_{it-1}$). In the absence
of movement, $\omega$ is exactly the probability of surviving from
year $t$ to $t-1$, and $\gamma$ is the per-capita birth rate. However,
in the more common scenario when immigration and emigration occur,
$\omega$ is the probability of surviving and not emigrating, and
$\gamma$ is the sum\hl{[I think!]} of the birth rate and the emigration rate,
i.e. the recruitment rate. \hl{[Feels a little repetitious with the material
before the equation, although I understand you're describing S and G
there and omega and gamma here.  In general, these sections
describing the DM models are clear and well written, but we could
possibly spend less time on them b/c it's already published.
Possibly it's better this way, though.]} \citet{dail_madsen:2011} proposed three
models for $\gamma$: the constant model,
$G_t \sim \mathrm{Pois}(\gamma)$ where recruitment does not depend on
$N_{it-1}$, and which simulates a ``propagule rain'' of new individuals; the
autoregressive model, $G_t \sim \mathrm{Pois}(\gamma(N_{it-1}))$, which
simulates geometric or density independent growth; and the
``no-trend'' model, $\gamma = (1-\omega)\Lambda$, which keeps
expected abundance constant over time. As before, covariates of
$\omega$ and $\gamma$ can be easily accomodated, for example using
logit- and log-linear models respectively.

In cases where no movement occurs, the DM model can directly estimate
demographich parameters, and this mechanistic description of
population change is much more appealing than classical
population growth models such as the logistic model. However, movement
complicates the interpretation of the the parameters, reducing the
connection to specific population processes. In such cases, the model
of dynamics shown in Eq~\ref{eq:Nt} can be replaced with standard
population models as described in Sect~\ref{sec:ext}. Irrespective of
the model for population dynamics, it is worthwhile to note that the
discrete distributions used in the DM model avoid the issues
associated with the Gaussian distribution used to model process
variation in standard state-space models.

>>>>

As with traditional state-space models, the DM model includes the
three conditionally related processes shown in Eq.~\ref{eq:ss2}, which
correspond to (1) initial abundance, i.e. the
abundance at site $i$ during the first primary period,
denoted (2) abundance at time $t$ (for $t>1$) which depends upon
abundance at $t-1$, and (3) the
detection process \citep{dail_madsen:2011}.
The first two processes describe the state process---the
variation in abundance in space and time. The third process simply
describes the the relationship between
abundance and the observed count data.

@@

In the original formulation, initial abundance was
modeled using either the Poisson or negative binomial
distributions. Hence, in the Poisson case, we have
\begin{equation}
  [N_{i1} | \Lambda] = \frac{\Lambda^{N_{i1}}e^{-\Lambda}}{N_{i1}!}.
  \label{eq:N1br}
\end{equation}
where $\Lambda_i$ is the expected abundance at site $i$ during year 1.
From hereout, we will reserve bracket notation for arbitrary
probability distributions, and for simplicity, use the equivalent notation $N_{i1}
\sim \text{Pois}(\Lambda_i)$ when referencing specific distributions.
%\begin{gather}
%N_{i1} \sim \mathrm{Pois}(\Lambda) \nonumber \\
%\text{or} \nonumber \\
%N_{i1} \sim \mathrm{NB}(\Lambda, \alpha)
%\label{eq:N1}
%\end{gather}
The Poisson distribution assumes that the mean of $N_{i1}$ is equal to its
variance, whereas the negative binomial distribution, $N_{i1} \sim
\text{NB}(\Lambda, \alpha)$, allows the
variance to be greater than the mean with the amount of overdispersion
determined by the parameter $\alpha$.

Regardless of the specified distribution, the model for initial
abundance has two distinguising features. First, it provides a
mechanism for characterizing spatial variation in abundance. For
instance, one might consider the influence of some environmental
covariate ($x_i$) on abundance using a log-linear
model such as $\log(\Lambda_i) = \beta^{\Lambda}_0 + \beta^{\Lambda}_1
x_{i}$. The second important point is that the spatial replicates resolve the
problem of parameter non-identifiability that are common to standard
state-space models because, as demonstrated by \citet{royle:2004biom},
process variation and observation error can be estimated from
spatially-replicated count data. Hence, the first component of the
model addresses both the issues of spatial inference and parameter
identifiability discussed previously.

@@

The DM model assumes that abundance in time $t$ is a function of
abundance in time $t-1$, i.e. abundance at each site evolves as a
first order Markovian process, although
higher order processes are also possible.  \citet{dail_madsen:2011}
considered several models to describe the temporal dynamics; however,
in each case they modeled $N_t$ as the sum of two random variables:
$S_{it}$, the number of indivduals surviving from $t-1$ and not
emigrating; and $G_{it}$ the number of new individuals entering the
population. Their most general model was
\begin{equation}
\left.\begin{aligned}
S_{it}|N_{it-1} &\sim \mathrm{Bin}(N_{it-1}, \omega) \\
G_{it}|N_{it-1} &\sim \mathrm{Pois}(\gamma(N_{it-1})) \\
N_{it} &= S_{it}+G_{it}
\end{aligned}\right\} \quad \text{for} \; t=2,\hdots,T
\label{eq:Nt}
\end{equation}
where $\omega$ is the apparent survival probability and $\gamma$ is
the recruitment rate (which can depend on $N_{it-1}$).
%%%% I agree with your comment, but I think we need to distinguish
%%%% between things like actual and apparent survival probs.
%\hl{[Feels a little repetitious with the material
%before the equation, although I understand you're describing S and G
%there and omega and gamma here.  In general, these sections
%describing the DM models are clear and well written, but we could
%possibly spend less time on them b/c it's already published.
%Possibly it's better this way, though.]}
\citet{dail_madsen:2011} proposed three
models for $\gamma$: the constant model,
$G_{it} \sim \mathrm{Pois}(\gamma)$ where recruitment does not depend on
$N_{it-1}$, and which simulates a ``propagule rain'' of new individuals; the
autoregressive model, $G_{it} \sim \mathrm{Pois}(\gamma(N_{it-1}))$, which
simulates geometric or density independent growth; and the
``no-trend'' model, $\gamma = (1-\omega)\Lambda$, which keeps
expected abundance constant over time. As before, covariates of
$\omega$ and $\gamma$ can be easily accomodated, for example using
logit- and log-linear models respectively.

In the absence
of movement, $\omega$ is exactly the probability of surviving from
year $t$ to $t-1$, and $\gamma$ is the per-capita birth rate. However,
in the more common scenario when immigration and emigration occur,
$\omega$ is the probability of surviving and not emigrating, and
$\gamma$ is the sum %\hl{[I think!]}
of the birth rate and the immigration rate,
i.e. the recruitment rate. In cases where no movement occurs, the DM model can directly estimate
demographich parameters, and this mechanistic description of
population change is much more appealing than classical
population growth models such as the logistic model. However, movement
complicates the interpretation of the the parameters, reducing the
connection to specific population processes. In such cases, the model
of dynamics shown in Eq~\ref{eq:Nt} can be replaced with standard
population models as described in Sect~\ref{sec:ext}. Irrespective of
the model for population dynamics, it is worthwhile to note that the
discrete distributions used in the DM model avoid the issues
associated with the Gaussian distribution used to model process
variation in standard state-space models.
